---
title: "Credit Card Fraud Detection"
author: "Maroofa Sheikh"
date: "2025-10-01"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
---

# 📌 Project Overview
This project focuses on **Credit Card Fraud Detection** using machine learning in R.  
The goal is to identify fraudulent transactions from a dataset containing anonymized credit card transaction details.

---

# 📊 Dataset
- **Source:** [Kaggle Credit Card Fraud Detection Dataset](https://www.kaggle.com/mlg-ulb/creditcardfraud)  
- **Size:** 284,807 transactions with 31 features  
- **Target Variable:** `Class` (1 = Fraud, 0 = Normal)

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Load Libraries
```{r}
library(dplyr)
library(ggplot2)
library(caret)
```

# Load Data
```{r}
creditcard <- read.csv("creditcard.csv")
```

# Check dataset structure
```{r}
str(creditcard)
summary(creditcard$Class)
```

# 📈 Exploratory Data Analysis (EDA)

# Distribution of classes
```{r}
table(creditcard$Class)

ggplot(creditcard, aes(x = factor(Class), fill = factor(Class))) +
  geom_bar() +
  labs(title = "Class Distribution", x = "Class (0 = Normal, 1 = Fraud)", y = "Count") +
  scale_fill_manual(values = c("0" = "skyblue", "1" = "red")) +
  theme_minimal()
```

# Summary statistics for numeric features
```{r}
summary(select(creditcard, -Class))
```

# Optional: Correlation heatmap for numeric features
# Select only numeric columns
```{r}
numeric_vars <- select(creditcard, where(is.numeric))
```

# Compute correlation matrix
```{r}
cor_matrix <- cor(numeric_vars)
```

# Install ggcorrplot if not installed


# Load plotting library
```{r}
library(ggcorrplot)
```

# Plot correlation heatmap
```{r}
ggcorrplot(cor_matrix, hc.order = TRUE, type = "lower", lab = FALSE,
           title = "Correlation Heatmap of Numeric Features")
```

# 🔧 Data Preprocessing

# Convert target variable to factor
```{r}
creditcard$Class <- as.factor(creditcard$Class)
```

# Check class distribution again
```{r}
table(creditcard$Class)
```

# Split data into training and testing sets (70%-30%)
```{r}
set.seed(123)  # for reproducibility
trainIndex <- createDataPartition(creditcard$Class, p = 0.7, list = FALSE)
trainData <- creditcard[trainIndex, ]
testData <- creditcard[-trainIndex, ]
```

# Confirm split
```{r}
table(trainData$Class)
table(testData$Class)
```

# 🤖 Model Training - Random Forest
```{r}
library(randomForest)
library(pROC)
```

# Train Random Forest model
```{r}
set.seed(123)
model_rf <- randomForest(Class ~ ., data = trainData, ntree = 100)
```

# Make predictions
```{r}
pred_rf <- predict(model_rf, testData)
pred_prob_rf <- predict(model_rf, testData, type = "prob")[,2]
```

# Confusion Matrix
```{r}
confusionMatrix(pred_rf, testData$Class, positive = "1")
```

# ROC Curve & AUC
```{r}
roc_rf <- roc(testData$Class, pred_prob_rf)
plot(roc_rf, col = "blue", main = "ROC Curve - Random Forest")
auc_value <- auc(roc_rf)
auc_value
```

# Feature Importance
```{r}
varImpPlot(model_rf, main = "Feature Importance - Random Forest")
```



# 📜 Conclusion

The Random Forest model was able to detect fraudulent transactions with high accuracy.  

**Key insights:**
- Fraud detection requires balancing **precision** and **recall**, as false negatives (missed frauds) are costly.  
- The model highlights important features that contribute most to detecting fraud.

**Future improvements:**
- Apply **SMOTE oversampling** or other techniques to handle class imbalance.  
- Compare multiple models (e.g., Logistic Regression, XGBoost, Gradient Boosting).  
- Fine-tune hyperparameters for better performance.  
- Deploy the model for real-time fraud detection in transaction systems.

